---
title: "Reap What You Sow"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../docs") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library("plotly", quietly = TRUE)
```


## Planting Strategy
Garlic is generally planted in the fall and harvested in the summer. In my garden near the [Denver metro](https://dug.org/how-to-grow-garlic/#:~:text=On%20the%20High%20Plains%2C%20depending,The%20stalk%20will%20break.), that means mid October to early to mid July. 

Several factors, not just clove weight, can impact growth: uneven soil/nutrient densities, irrigation/drainage, and wind/sun. 

The location that each clove was planted in can serve as a proxy for those factors. The plants on the perimeter of the raised bed will experience more drainage. Certain sides of the box may get more or less light/shading as the rest of the garden grows.

I don't want my planting strategy to confound these factors with weight. If I plant the large cloves in one spot and the small ones in another, how will I know if the size or the location is causing the difference?

I used R to help me randomize the planting locations. Assigned each clove an ID (number 1 through 36) then had R randomize the order.

``` {r}
clove_weight <- c(8, 17, 14, 20, 13, 13, 11, 15, 11,
                 10, 19, 19, 19, 13, 17, 13, 15, 11,
                 12, 19, 14, 13, 21, 8, 5, 13, 12, 13,
                 17, 9, 8, 17, 9, 20, 7, 13)
id <- 1:36
randomized_example <- sample(id)
randomized_example
planting_order <- c(23, 18, 2, 36, 3, 27, 12, 11, 1,
                    35, 6, 21, 9, 24, 28, 22, 25, 31,
                    33, 8, 4, 15, 14, 26, 5, 16, 13,
                    7, 17, 10, 19, 20, 32, 29, 34, 30)
```



The `randomized_example` above is different every time you load this page. (As it should be. It's random!) 

The randomization I used is recorded as `planting_order`. Clove #`r planting_order[1]` is at the North East corner. Clove #`r planting_order[length(planting_order)]` is in the South West corner of the raised bed.

``` {r}
dimensions <- c(4, 9) #East to West, North to South
row_from_east <- c(rep(c(1:dimensions[1]), each = dimensions[2]))
row_from_north <- c(rep(1:dimensions[2], times = dimensions[1]))

garlic <- data.frame(id = planting_order,
                     from_east = row_from_east,
                     from_north = row_from_north,
                     year = rep(2024, times = length(id)))
```

I'd like to merge the position information in the `garlic` dataframe with the weights in the `clove_weight` vector.

Since `garlic` is randomly ordered, we need to re-order it by `id` to match the order of the weigths in the `clove_weight` vector.

``` {r}
head(garlic) #randomized IDs
garlic <- garlic[order(garlic$id), ]
head(garlic) #sorted by IDs

#create new column using data from vector
garlic$weight <- clove_weight

#renumber the rowname numbers
rownames(garlic) <- 1:nrow(garlic)
head(garlic)
```

Here's a diagram of where each clove was planted.
``` {r, echo=FALSE}
p <- plot_ly(data = garlic, type = "scatter", mode = "markers") %>%
  add_markers(x = ~from_east, y = ~from_north,
              marker = list(color = 'rgb(17, 157, 255)',
                            size = ~weight * 2,
                            line = list(color = 'black',
                                        width = 2)
      ),
              text = ~paste('<br> ID: ', id,
                            '<br> Weight: ', weight, ' grams'),
              hoverinfo = 'text',
              showlegend = F ) %>%
  layout(xaxis = list(autorange = "reversed"),
         yaxis = list(autorange = "reversed"))
p
```

